---
title: "forcedlabor: forced labor model package"
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

Updated: `r Sys.Date()`

# Overview

`forcedlabor` is an R package containing functions to identify forced labor
on fishing vessels via a machine learning algorithm trained on a combination of
known forced labor cases and AIS data predictors.

# Installation

You can install the most recent version of `forcedlabor` using:



```{r eval = FALSE}

# Check/install remotes
if (!require("remotes"))
  install.packages("remotes")

# Install the forcedlabor package

remotes::install_github("GlobalFishingWatch/forcedlabor",
                        dependencies = TRUE)
```

# Using the package

An example of how to use it is in a paper repo (soon to come)


# Structure of the repository


* **R:**            R functions of the package
* **man:**          Standard documentation files for R functions
* **renv:**         Contains files to activate the R environment with specific
                    settings
* **.Rbuildignore** Contains names of files that should be ignored to build
                    the R package
* **.Rprofile**     Used in the R session to get the R environment of this
                    package
* **.gitignore**	  Files to ignore by version control.
* **.pre-commit-config.yaml** Style linting using `pre-commit library.
* **DESCRIPTION**   It stores the metadata of the package
* **NAMESPACE**     It contains information of important names in the package
                    to provide space for them and make the package self
                    contained
* **README.md**		  Top-level README on how to use this repo
* **README.Rmd**    Rmarkdown file that generates README.md
* **forcedlabor.Rproj** Rproj file
* **renv.lock**     It contains the information on the packages used in the
                    project.
