---
title: "forcedlabor: forced labor model package"
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

Updated: `r Sys.Date()`

# Overview

`forcedlabor` is an R package containing the codes to identify forced labor
on fishing vessels via a machine learning algorithm trained on a combination of
known forced labor cases and AIS data predictors.

# Installation

This R package is hosted on a private GitHub repository and can be installed
using the `devtools` package. Run the following code to check if you already
have `devtools` and install it if not.


``` r
# Check/install devtools
if (!require("devtools"))
  install.packages("devtools")
```

To connect to GitHub, under the hood `devtools` calls the GitHub API. This means
that you’ll need to have a personal access token (PAT). Get a PAT
[here](https://github.com/settings/tokens) and make sure the “repo” scope is
included.
Save your PAT as an R environment variable (variables that should be available
to R processes) by running `usethis::edit_r_environ()`, adding `GITHUB_PAT =
"your PAT"`, and saving the updated `.Renviron` file. You might need to restart
the R session.

Finally, after saving your GitHub PAT to your R environment, install the
`forcedlabor` package using `devtools`.

```{r eval = FALSE}
# Install fishwatchr
devtools::install_github("GlobalFishingWatch/forcedlabor")
```

# Using the package

An example of how to use it is in
[`/scripts/model_run.R`](https://github.com/GlobalFishingWatch/forcedlabor/blob/main/scripts/model_run.R)

# Structure of the repository

Names in **bold** correspond to folders and files that are currently used.

* **R:**            R functions of the package
* **data:**			    Data files. For now, it contains a couple of output data
                    objects.
* docs:			        Additional documentation (empty for now)
* **man:**          Standard documentation files for R functions
* models:		        Trained models, model summaries, etc. (empty for now)
* notebooks: 	      Jupyter notebooks and R markdown (empty for now)
* outputs:		      Models results, static reports, etc.
  * figures:	      Figures (empty for now)
* queries:		      SQL files (empty for now)
* **renv:**         Contains files to activate the R environment with specific
                    settings
* **scripts:**		  Regular python and R files. Contains the script to run the
                    model.
* **.Rbuildignore** Contains names of files that should be ignored to build
                    the R package
* **.Rprofile**     Used in the R session to get the R environment of this
                    package
* **.gitignore**	  Files to ignore by version control.
* **.pre-commit-config.yaml** Style linting using `pre-commit library.
* **DESCRIPTION**   It stores the metadata of the package
* **NAMESPACE**     It contains information of important names in the package
                    to provide space for them and make the package self
                    contained
* **README.md**		  Top-level README on how to use this repo
* **README.Rmd**    Rmarkdown file that generates README.md
* **forcedlabor.Rproj** Rproj file
* **renv.lock**     It contains the information on the packages used in the
                    project.
