---
title: "Saving and loading models"
author: "AST"
date: "2026-04-24"
output: html_document
---

# Alternative workflows with `ml_load()` and `ml_predict()`

Function `ml_train()` can train the model and create predictions both for the training data and for a new dataset. 
When new cases arrive (a new `holdout_df` that needs predictions to be generated), an alternative workflow can be:

- saving the models to disk using the `save_dir` parameter of `ml_train()`
- load the previously fitted models to the R session using `ml_load()`
- create a new set of predictions using function `ml_predict()`

### Tuning models

The example provided had already fixed hyperparameters to train the PU model. 

To arrive to a combination of parameters 

### Saving models on disk 

To save the model objects on disk the user can use parameter `save_dir` to point the destination folder.
Note that this process will take a while since the trained modes are quite heavy. 

```{r ml_train_save}
tictoc::tic()
trained_models <- purrr::map(.x = cv_df,
                             ~ml_train(rf_spec = rf_setup$rf_spec,
                                       holdout = holdout_df,
                                       save_dir ="test"))

tictoc::toc()

```

```{r}
knitr::knit_exit()
```

### Loading previously fitted models 

Previously trained models can be loaded from your local disk using `ml_load()` based on your cross-validation setup (output of `cv_setup()`).


```{r load_models, eval = FALSE}
loaded_models <- ml_load(cv_setup = cv_df,
                         save_dir = "./vignettes/test")
```

### Predicting over new data 

Fitted models can be used to predict over new data using `ml_predict()`. 

```{r rf_predictions , eval = FALSE}
predictions <- ml_predict(trained_models = trained_models)
```

`predictions` contains the same columns as `trained_models$train_probabilities`, but `holdout` takes value of 1, as predictions were generated over an external dataset not used for training.
