---
title: "Forced labor risk workflow on a sample dataset"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Forced labor risk workflow on a sample dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}"
execute:
  project
  
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

devtools::load_all()
```

## Overview

`forcedlabor` is an R package develop for training a positive–unlabeled (PU) 
model to detect fishing vessels exhibiting behaviors or characteristics 
consistent with forced-labor offenders. 

The model characteristics, particularly the PU approach, obey to the nature of 
the available data on forced labor:

- We have __positive cases__ of forced labor, which refer to vessels reported to 
have engaged in forced labor. The model uses _vessel-year_ as a unit for forced 
labor cases. 
- While there are confirmed reports of certain vessels engaged in forced labor, 
a confirmation of the absence of forced labor is extremely difficult to find. 
The immense majority of vessels are not inspected for labor exploitation, and 
even those that are, are usually inspected at port. A certification in port 
is not 100\% guarantee of the absence of forced labor at sea. 
So instead of having negative cases to train the model along with the positive 
ones, the model takes __unlabeled cases__--with an unknown mix of positive and
negative cases. __Unlabeled cases__ are thus vessel-year combinations for which we have 
no information (no evidence of being positive but also no evidence of them
being negative). 
- We had access to a list of vessels certified as compliant with the ILO C188
requirements for work and living conditions. These certifications were based on 
documents and inspections at port. They can be seen as __negative cases__, but 
for the reasons mentioned above, they are not used for training the model. We 
rather generate predictions for them after the training, expecting them to be 
classified as negative. 

## The dataset in this package

The dataset, `fl_training`, comes from a real dataset of positive and unlabeled
cases, with one row per vessel-year and columns corresponding to vessel identity,
the features of the model based on AIS data (and briefly described in a section 
below), and columns to identify the positive cases (`known_offender` column), 
negative cases (`known_non_offender` column), and unlabeled cases (`0` values in 
`known_offender` and in `known_non_offender`). 
The original dataset contained less than 100 positive cases and almost
a million vessel-years of unlabeled data. The vessel gears included are
trawlers, longlines and purse seines. To generate `fl_training`, we 
anonymized the vessels and randomly sampled 10,000 vessel-years, so that there
could be a manageable sample to run the functions in the package in a local 
computer. This sample does not contained real negative cases (the 
C188-certified vessels), as they were obtained under a non-disclosure 
agreement. If further negative cases are made available and it is possible to 
share them publicly, we could add them to the holdout dataset of the package in
the future. 

## Model features

The features used for classification correspond to vessel characteristics and variables describing vessel activity at sea that can be indicators of forced labor. 

The features for this models were computed from Automatic Identification System (AIS) data processed by Global Fishing Watch. 

The rationale behind why we selected these features is provided below.

### Vessel characteristics

- Gear type shape expected movement.
- Length, tonnage and engine power may enable longer, farther operations—factors 
associated with elevated forced-labor risk.

### Time at sea

Long voyages and infrequent port allow vessels to avoid inspections and limit
crew opportunities to report or exit abusive conditions.
Excessive overtime is a common indicator of abuse. The following variables were
included to capture unusually long daily or yearly effort:

- number of AIS positions
- number of hours at sea
- number of voyages
- average voyage duration (in hours)
- total number of fishing hours and average daily fishing hours

### Distance from shore and port

Operating far from shore may mean greater isolation and reduced oversight. The model included: 
 
- maximum distance from port (km)
- maximum distance from shore (km)

### Multijurisdictional issues

Time in foreign EEZs or on the High Seas, and foreign post visits, increase
the jurisdictional complexity to address forced labor cases. The model included: 
- number of fishing hours in foreign EEZs
- number of fishing hours in the High Seas
- number of foreign port visits

### Transshipment, encounters, and loitering

Encounter and loitering (stationnary behavior at sea that may indicate a  
potential encounter with a non-AIS-broadcasting vessel) patterns can indicate
transshipment events. 
Transhipments can prolong the time at sea and enable crew transfers, both of
which are linked to higher risk of forced labor.

- number of encounters with other vessels at sea
- number of encounters with vessels reported to have engaged in forced labor,
- average encounter duration (hours)
- number of loitering events
- average loitering duration (hours).

### AIS disabling events-"gaps"

Intentional AIS disabling can obscure activity and mask behaviors such as 
transshipment or other evasive practices.



```{r setup, eval = F}
library(forcedlabor)
```

## Model workflow 

The `forcedlabor` package allows to execute the workflow described in [@joo_2023]. This model first fits a series of Random Forests, using the `ranger` engine implementation and a `tidymodels` framework.

- Function `fl_rfsetup()` configures the model features, response variables, and the parameters required to configure the Random Forests (number of trees, `mtry`, minimum n per tree, regularization factor). 
- Function `fl_cvsetup()` takes the output of `fl_rfsetup()` and configures the number of folds, bags, seeds and downsampling ratio required to run the crossvalidations.
- Function `fl_train` fits the Random Forests for each combination of bag, fold and seed using the `recipe` configuration from `fl_cvsetup()`, creating a series of predictions for the training data, and optionally, for the holdout dataset. The function is written to generate the output for a single Random Forest (all the folds), and this implementations allows for a run of all models in sequence (using a `for` loop or `purrr::map()`) or parallel (like `furrr::future_map()`). 
- Function `fl_classification()` implements the DEDPUL algorithm, calculating a calibrated threshold `t` and classifying the random forest mean scores against `t` to classify each prediction as `positive` or `negative`. 
The calculation of confidence values for each case can also be done. 
- Function `fl_perf_metrics()` calculates recall and specificity (if a `holdout` dataset is provided). 


```{r workflow, echo=FALSE, out.width="100%"}
here <- here::here()
img <- (glue::glue(here, "/vignettes/fig01_model_workflow.png"))
knitr::include_graphics(img, rel_path = FALSE)

```


## Train and test data

The model starts with loading the required training data (`training_df`).
This table contains AIS-derived data at a yearly scale for positive vessels 
(`known_offender = 1`) and unlabeled vessels (`known_offender = 0`). 

We will split this example dataset into two dataframes: `training_df` will be used to train the model and `holdout_df` will be used to evaluate the model.

- The original model is not trained on negative cases (`known_non_offender = 1`) 
- If negative data is available, it can be used in the `holdout_df` to 
calculate specificity (but `holdout_df` is not used in training)
- If non-offender data is not available, `holdout_df` should be set to `NULL` 
- Note that when `holdout_df` = `NULL`, specificity cannot be computed

```{r load_data}
data("fl_training")

set.seed(101)
rows_pred <- sample(seq_len(nrow(fl_training)), size = 1000)
training_df <- fl_training[-rows_pred, ]
holdout_df <- fl_training[rows_pred, ]
```


To reduce the risk of data leakage, we keep data from the same source (in our
case from the same forced labor report) in the same group.
This means that vessels with the same `source_id_number` will either all be
in training or all in test.
We do this as two vessels from the same report are likely to have similar
behaviors and characteristics, and so when using one of them in the 
training set, the model would likely
perform well on the other one, an artificial effect of using related cases to 
train and test.

```{r cv_splits}
num_folds <- 5
num_bags <- 10
down_sample_ratio <- 1
num_common_seeds <- 3
```


## Model configuration using `fl_rfsetup()` and `fl_cvsetup()`

There are two aspect to define as part of our random forest model, the model recipe, and the model specifications.

#### Writing the recipe

Here we set the roles of the different variables that will be used. 

**Setting the roles of the variables used**

- `known_offender` is set as the **outcome** i.e. the target we predict
- `indID` is set as an "id" value and is not used as a predictor
- `known_non_offender` is not used as a predictor (role "dont_use")
    - if available, use `known_non_offender`  in a holdout `holdout_df` to 
      compute specificity later
- `source_id_number` is set as "control"
- `flag_region` is not used as a predictor (role "dont_use")
      - it is reserved for post-hoc fairness checks 
- **Predictors**: All remaining variables in `training_df` are treated as 
predictors

**Variable preprocessing**

- We use `step_nzv()` to retain predictors with non-zero variance
  - This step diagnoses predictors that have one unique value 
  (i.e. are zero variance predictors) or predictors that have both of the 
  following characteristics:
    1. they have very few unique values relative to the number of samples
    2. the ratio of the frequency of the most common value to the frequency of 
  the second most common value is large.
- Using `step_corr` we only keep predictors with bivariate correlations smaller 
than 0.75.

#### Specifying the model

Next, we define the classifier. 
We use a Random Forest via `parsnip` with the `ranger` engine.

- The type of model used is a random forest `rand_forest()`
- `trees = 500` sets the number of trees contained in the ensemble is set to 
500
  - The larger the number of trees, the more stable the predictions - but with
  a higher computational cost
- `mtry = 1` sets the number of features that will be sampled at each
split
  -  `mtry` must be ≤ the number of predictors in the dataset
  - Lower values encourage diversity between trees (higher variance, 
  less correlation)
  - Higher values let the model consider more predictors, potentially improving 
  accuracy but also with a risk of overfitting
- `min_n = 15` sets the minimum node size (i.e. the number of vessel-years that 
a final node should have)
  - Larger values create less complex trees, and may result in underfitting
  - Smaller values create more complex trees, and may result in overfitting
- `set_mode("classification")` the prediction outcome mode is set as 
"classification" 
- `set_engine("ranger", regularization.factor = 0.5)` 
- `set_engine()` 
  - `ranger` is particularly useful when dealing with data with a high number of
  features as it provides memory-efficient yet quick implementation
- `regularization.factor = 0.5` sets the regularization factor works by 
penalizing new variables by multiplying the splitting criterion
by a factor
  - As a result, the tree is biased to keep splitting on features it already 
  trusts, rather than constantly introducing fresh ones
    - `f = 0.5`: half the gain for “new” variables, forcing the forest to only 
    bring in a new predictor if it improves impurity by at least twice as much
    the default would require

```{r fl_rfsetup}
features_model_characteristics <- c("gear",
                                    "engine_power_kw",
                                    "tonnage_gt",
                                    "length_m")
features_model_movement <- c("position_messages",
                             "hours",
                             "fishing_hours",
                             "average_daily_fishing_hours",
                             "fishing_hours_foreign_eez",
                             "fishing_hours_high_seas",
                             "max_distance_from_shore_km",
                             "max_distance_from_port_km",
                             "number_encounters",
                             "number_forced_labor_encounters",
                             "average_encounter_duration_hours",
                             "gaps_12_hours",
                             "average_off_distance_from_port_km",
                             "average_off_distance_from_shore_km",
                             "average_gap_days",
                             "average_gap_km",
                             "number_foreign_port_visits",
                             "number_loitering_events",
                             "average_loitering_duration_hours",
                             "average_voyage_duration_hours",
                             "number_voyages")
rf_setup <- fl_rfsetup(training_data = training_df,
                       y = "known_offender",
                       x = c(features_model_characteristics,
                             features_model_movement),
                       id = "indID",
                       dont_use = c("flag_region", "known_non_offender"),
                       control = "source_id_number",
                       corr_threshold = 0.75,
                       rf_trees = 500,
                       rf_mtry = 5,
                       rf_min_n = 10,
                       rf_reg_factor = 1)
```

### Function `fl_cvsetup()`

```{r fl_cvsetup}

tictoc::tic()
cv_df <- fl_cvsetup(training_data = training_df,
                    num_folds = num_folds,
                    num_bags = num_bags,
                    num_seeds = num_common_seeds,
                    fl_rec = rf_setup$rf_recipe,
                    rf_spec = rf_setup$rf_spec,
                    down_sample_ratio = down_sample_ratio)
tictoc::toc()
```


## Training the models: function `fl_train()`

Once the training data has been prepped and the model recipe and specifications
set, we can now use function `fl_train()`to fit models and generate scores. 

Function `fl_train()` fits a Random Forest for each fold x bag x seed using
the previous recipe, specification steps and cross-validation setup
- It applies a downsampling ratio of 1:1 to the training set
- This step returns list of  `train_probabilities` a table with probabilities for
assessment rows and the `fitted_models`. If a directory is specified, fitted models are saved accordingly. 
- `fl_train()` also pools scores across all seeds and bags

The function is written to receive a single element of the output of `fl_cvsetup()` (`cv_df` in this example). So a single run (with 5 folds) 
could be generated like this:

```{r, eval = FALSE}
tictoc::tic()
one_trained_model <- fl_train(cv_setup = cv_df[[1]],
                              rf_spec = rf_setup$rf_spec,
                              holdout = holdout_df)
tictoc::toc()
```

This allows us to loop effectively across all seeds and bags, using `for` loops or `purrr::map()`, and parallelize in remote machines. 
In this example, let's use `purrr::map()`.

Note that the use of `holdout` is optional. If omitted, `fl_train()` will produce predictions for the training data only, if added, it will create predictions for new data, not used in model training.

```{r fl_train}
tictoc::tic()
trained_models <- purrr::map(.x = cv_df,
                             ~fl_train(rf_spec = rf_setup$rf_spec,
                                       holdout = holdout_df))

tictoc::toc()

```

The output of `fl_train()` is another list of length `bags x seeds` (in this case, `r num_common_seeds * num_bags`).
Each element of this list contains two tibble objects with the results for a single `fold`: `train_probabilities` and `pred_probabilities` __if `holdout` was included__.

Both tibbles include the following columns: 

- `.pred_1`: the model’s predicted probability
- `indID`: a unique vessel identifier
- `known offender`: if the vessel is positive
- `known non offender`: if the vessel is negative (unavailable in this dataset)
- `holdout`:  0 in the case of `train_probabilities` and 1 in the case of `pred_probabilities`, if available (i.e., if `holdout` was passed to the function).

Let's extract both lists and create a single dataframe for all

```{r train_proba_output}

train_probabilities_all <- purrr::map(trained_models, "train_probabilities") |> bind_rows()
pred_probabilities_all <- purrr::map(trained_models, "pred_probabilities") |> bind_rows()
prob_all <- bind_rows(pred_probabilities_all,
                      train_probabilities_all)
head(prob_all)
```

### Function `fl_classification()`: density-based algorithm for classification threshold

`fl_classification()` aggregates all random forest predictions for each vessel-year, 
estimates $\alpha$ (upper bound on positives in the unlabeled set) based on $\beta$ probabilities, 
and chooses a calibrated threshold `t` so the share of 
predicted positives ≈ $\alpha$.

See [@ivanov_2020] for more details on the DEDPUL algorithm.

Before we continue, it may be worth mentioning parallelization. `fl_train()` can be parallelized if needed, but the number and size of the models allows for running them in sequence, but `fl_classification()` is generating confidence values across the whole unlabelled dataset and for a sequence of values. 
In Joo et al 2026, this meant 722,000 unlabelled cases, so the function is set to run in parallel using `furrr::future_pmap()` function internally. 

To set up a parallelization section to run `fl_classification()`, we can load and configure a `future` plan:


```{r classif_res, warning=FALSE}
tictoc::tic()
classif_res <- fl_classification(data = prob_all,
                                 steps = 1000,
                                 plotting = FALSE,
                                 filepath = NULL,
                                 threshold = seq(0, .99, by = 0.01),
                                 eps = 0.1,
                                 confidence_levels = TRUE
                                 )
tictoc::toc()
plan(sequential)
```

- Within `fl_classification()` we apply the calibrated
threshold `t` to each vessel’s mean score:  
  `pred_class = 1` if `pred_mean > t`, else `0`. This produces the binary prediction (positive or negative, stored in column `pred_class`). 
- The `confidence` value indicates how consistently a vessel’s
score falls on the same side of `t` across seeds × bags. Values near 1
indicate a very stable classification for that vessel-year (higher
confidence in the classification).

In the case of this sample dataset, 8088 vessel-years were classified as negative and 1984 were classified as positive for forced labor risk. 

```{r classif_res_pred_conf}
classif_res$pred_conf |> count(pred_class)
```

## Additional steps: Performance and Fairness


### Performance: Recall and specificity (using `fl_perf_metrics()`)

The performance of the model is evaluated through:

- `recall`: the proportion of positive cases predicted by the model
as positive
- `specificity`: if a `holdout_df` has been provided, the
proportion of  negative cases predicted by the model as negatives

```{r perf_metrics}

tictoc::tic()
perf_metrics <- fl_perf_metrics(data = classif_res$pred_conf)
tictoc::toc()

print(perf_metrics)


```


### 13. Fairness tests

Statistical non-discrimination or fairness measures aim to assess the absence or
degree of discrimination of given groups regarding classification. 

We evaluated the model’s fairness as its capacity to correctly identify reported 
positive cases by flag region and fishing gear.

The first example below shows how to assess fairness based on 'gear'

- The fairness test assesses whether the model is equally successful at 
classifying offenders as positive (recall) when considering different gears
  - Gears groups are disbalanced in the training data, which can create different results for different gears


```{r fairness_gear}

# Recall by gear

gear_lookup <- training_df  |>
  distinct(indID, gear)

gear_lookup |> count(gear)

gear_lookup |> 
left_join(classif_res$pred_conf) |>
 group_by(gear) |> 
 mutate(
   known_offender = factor(as.character(known_offender), levels = c("0","1")),
   pred_class     = factor(as.character(pred_class),     levels = c("0","1"))
 ) |> 
 yardstick::recall(truth = known_offender,
        estimate = pred_class, event_level = "second")  |> 
 select(gear, .estimate)
```




# References


