---
title: "Using the package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using_the_package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)

devtools::load_all()
```



Let's start by calling the package and the sample data
```{r setup}
library(forcedlabor)

```

##### Training data

The training data consists of vessel data at a yearly scale. This includes:

- known_offender: if 1, then the vessel has a reported case of forced labor
in that year. If 0, the vessel has no reported cases of forced labor in that 
year
- known_non_offender: if 1, then the vessel passed a labor-related inspection
in that year. If 0, the vessel did not pass any labor-related inspections in 
that year
- source_id_number: one per forced labor report source
- indID: an individual ID per vessel per year
- flag_region: used later for fairness tests 

vessel characteristics:

- gear, **ais_type**, **foc**, engine_power_kw, tonnage_gt, length_m

vessel behavior (based on AIS)

- position_messages, hours, fishing_hours, average_daily_fishing_hours, 
fishing_hours_foreign_eez, fishing_hours_high_seas, **distance_traveled_km**,
max_distance_from_shore_km, max_distance_from_port_km, number_encounters,
number_forced_labor_encounters, average_encounter_duration_hours, 
gaps_12_hours, average_off_distance_from_port_km, 
average_off_distance_from_shore_km, average_gap_days, average_gap_km,
number_foreign_port_visits, **number_poc_port_visits**, number_loitering_events,
average_loitering_duration_hours, average_voyage_duration_hours,
number_voyages



```{r}

data("new_training_sample")

new_training_data$known_non_offender <- as.factor(new_training_data$known_non_offender)

levels(new_training_data$known_non_offender) <- 
  c(levels(new_training_data$known_non_offender),1)


```

##### Writing the recipe

`fl_rec` writes the recipe for the model

- "known_offender" is set as the 'outcome' (what we are trying to predict)
- "indID" is used as an identifier, and should not be used as a predictor
- We remove "known_non_offender"
  - as we cannot say with high confident that vessels did not engage in forced 
  labor, so instead we are using a Positive-Unlabeled approach using 'positive' 
  and 'unlabeled' cases for training and testing
  - "known_non_offender" cases can be used later as 'expert-judgement-based
  validation' to compute specificity 
- We set "source_id_number" as a 'control'
  - To reduce the risk of data leaking, we kept data from the same source 
  in the same group, i.e. either all in training or all in test.
  - Two vessels that were mentioned in the same forced labor report are likely
  to show similar behaviors and may be related to each other, so when using one 
  of them in the training set, the model would likely perform well on the other 
  one, an artificial effect of using related cases to train and test.
- We remove "flag_region" as this is not used for predictions, but is
used later to assess fairness
- Using `step_nzv` we only keep predictors with variances computationally 
larger than zero.
  - This step diagnoses predictors that have one unique value 
  (i.e. are zero variance predictors) or predictors that have both of the 
  following characteristics:
  1. they have very few unique values relative to the number of samples
  2. the ratio of the frequency of the most common value to the frequency of 
  the second most common value is large.
- Using `step_corr` we only keep predictors with bivariate correlations smaller 
than 0.75.

-  **review** : "ais_type", "foc", "number_poc_port_visits", 
"distance_traveled_km" are all removed


```{r fl_rec}

################## writing the recipe ##########################################


fl_rec <- recipes::recipe(known_offender ~ .,
                          # modeling known_offender using everything else
                          data = new_training_data) |>
  # indID is just an identifier, and should not be used as a predictor
  recipes::update_role(indID,
                       new_role = "id") |>
  # Excluding these variables 
  recipes::update_role(flag_region, known_non_offender,
                       ais_type, foc,
                       number_poc_port_visits,
                       distance_traveled_km,
                       new_role = "dont_use")  |>
  # and some others will be useful for preventing data leakage
  recipes::update_role(source_id_number, new_role = "control")  |>
  # Remove near-zero variance numeric predictors
  recipes::step_nzv(recipes::all_predictors())  |>
  # almost zero variance removal (I don't think we have those though)
  # Remove numeric predictors that have correlation greater the 75%
  recipes::step_corr(recipes::all_numeric(), threshold = 0.75)

```


##### Specifying the model

`rf_spec` creates a random forest model specification using `parsnip`

- The type of model used in a random forest `rand_forest()`
- `trees = 500` sets the number of trees contained in the ensemble is set to 
500
- `mtry = 1` sets the number of features that will be sampled at each
split
- `min_n = 15` sets the min. node size (i.e. the number of vessel-years that a
final node should have)
- `set_mode("classification")`- the prediction outcome mode is set as 
"classification" (vs "unknown", "regression" or "censored regression")
- `set_engine("ranger", regularization.factor = 0.5)` sets the regularization
factor works by penalizing new variables by multiplying the splitting criterion
by a factor


```{r fl_spec}
######### specifying the model #################################################

# RF with hyperparameters based on sensitivity_hyper.r results
rf_spec <-
  # type of model # if no tuning # rand_forest()
  parsnip::rand_forest(trees = 500,
                       # We will tune these two hyperparameters
                       mtry = 1,
                       min_n = 15) |>
  # mode
  parsnip::set_mode("classification") |>
  # engine/package
  parsnip::set_engine("ranger", regularization.factor = 0.5)


```

##### Training and testing scheme

Defining some parameter values based on sensitivity results

- `num_folds <- 5` splits the data into 5 groups 
  - The data (positives and unlabeled) are split into five groups of roughly 
  equal size.
- `num_bags` sets the number of bags to run as 5
- `down_sample_ratio` downsampling ratio set to 1:1
  - To reduce the weight of the unlabeled cases in the model, we randomly 
  downsampled them in the training set with a 1-1 ratio, i.e. the number of 
  positive and unlabeled cases used for training would be equal.
- `num_common_seeds <- 3` sets 3 random seeds 
  - Setting up common seeds to use anywhere that uses random numbers.
- `cv_splits_all` checks that we don't have any splitting across source_id 
for the analysis and assessment datasets.

- `ml_train_predict` is used to train the RF models for each bag seed. 
For each bag seed, it generates a set of model hyperparameters configurations, 
fits a downsampled set of data to each model for each bag and each fold, and 
predicts over its corresponding assessment set (for each fold). 

- It produces `train_pred_proba` which is a data frame of trained
cross-validated datasets with several bags. 


```{r train_pred_proba}
########### training and testing scheme ########################################

## defining some parameter values based on sensitivity results
num_folds <- 5
num_bags <- 5 #10
down_sample_ratio <- 1 # downsampling ratio
# Set common seed to use anywhere that uses random numbers
num_common_seeds <- 3 #3
common_seed_tibble <- tibble::tibble(common_seed =
                                       seq(1:num_common_seeds) * 101)


# Run all common_seeds
bag_runs <- common_seed_tibble |>
  # For each common_seed, run all bags
  tidyr::crossing(tibble::tibble(bag = seq(num_bags))) |>
  # Will use different random seeds when implementing recipes for each bag
  dplyr::mutate(recipe_seed = dplyr::row_number() * common_seed) |>
  # counter
  dplyr::mutate(counter = dplyr::row_number())

## parallelization strategy
parallel_plan <- "multisession" # multisession if running from RStudio, or
# multicore if from Linux, Mac and plain R, or
# psock if multisession is not working well and you need to try something else
if (parallel_plan == "multisession"){
  utils::globalVariables("multisession")
}
free_cores <- 4 # add more if you need to do many things at the same time
oopts <- options(future.globals.maxSize = 10200*1024^2)  ## 15 GB
# options(future.globals.maxSize = 50 * 1024 ^ 3)

## CROSS VALIDATION ##
# Ensure there is no splitting across source_id across analysis and assessment
# data sets.  Need to make separate splits for each seed.
cv_splits_all <- common_seed_tibble |>
  dplyr::mutate(cv_splits = purrr::map(common_seed, function(x) {
    set.seed(x)
    rsample::group_vfold_cv(new_training_data,
                            group = source_id_number,
                            v = num_folds)
  }))


### training and prediction (scores)

tictoc::tic()
train_pred_proba <- forcedlabor::ml_train_predict(fl_rec = fl_rec,
                                             rf_spec = rf_spec,
                                             cv_splits_all = cv_splits_all,
                                             bag_runs = bag_runs,
                                             down_sample_ratio = down_sample_ratio,
                                             parallel_plan = parallel_plan,
                                             free_cores = free_cores,
                                             prediction_df = NULL)
tictoc::toc()

# 561.196 sec elapsed
# 1015.74 sec elapsed
# 1310.657 sec elapsed


```

##### Classification with dedpul

`ml_classification` computes binary classification via DEDPUL.

  -  For each vessel-year, it computes a binary classification, 0 non offender and 1 offender. 
  - It is based on the DEDPUL algorithm.

```{r classif_res}
####### Classification with dedpul ########################################
# alpha value will be printed

# I need to reinstall test branch of forcedlabor
tictoc::tic()
classif_res <- forcedlabor::ml_classification(data = train_pred_proba,
                                              steps = 1000, plotting = FALSE,
                                              filepath = NULL,
                                              threshold = seq(0, .99, by = 0.01),
                                              eps = 0.01,
                                              parallel_plan = parallel_plan,
                                              free_cores = free_cores)
tictoc::toc()

```

##### Recall and specificity


`ml_perf_metrics` computes recall for assessment sets and specificity for 
holdout non offenders

  - Two performance metrics are computed:
    1. recall, for assessment sets (in model versions that did not use them 
  for training), 
    2. specificity for holdout non offenders (in the same year of the
    certification/inspection - if done at the end of the year)

```{r perf_metrics}
########### Recall and specificity #################################################

tictoc::tic()
perf_metrics <- forcedlabor::ml_perf_metrics(data = classif_res$pred_conf)
tictoc::toc()

```
