---
title: "Forced labor risk workflow on a sample dataset"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Forced labor risk workflow on a sample dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}"
execute:
  project
  
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

devtools::load_all()
```

## Overview

`forcedlabor` is an R package developed for training a positive–unlabeled (PU) 
model to detect fishing vessels exhibiting behaviors or characteristics 
consistent with forced-labor offenders. 

The model characteristics, particularly the PU approach, obey to the nature of 
the available data on forced labor:

- We have __positive cases__ of forced labor, which refer to vessels reported to 
have engaged in forced labor. The model uses _vessel-year_ as a unit for forced 
labor cases. 
- While there are confirmed reports of certain vessels engaged in forced labor, 
a confirmation of the absence of forced labor is extremely difficult to find. 
The immense majority of vessels are not inspected for labor exploitation, and 
even those that are, are usually inspected at port. A certification in port 
is not 100\% guarantee of the absence of forced labor at sea. 
So instead of having negative cases to train the model along with the positive 
ones, the model takes __unlabeled cases__--with an unknown mix of positive and
negative cases. Unlabeled cases are thus vessel-year combinations for which we have 
no information (no evidence of being positive but also no evidence of them
being negative). 
- We had access to a list of vessels certified as compliant with the ILO C188
requirements for work and living conditions. These certifications were based on 
documents and inspections at port. They can be seen as __negative cases__, but 
for the reasons mentioned above, they are not used for training the model. We 
rather generate predictions for them after the training, expecting them to be 
classified as negative. 

## The dataset in this package

The dataset, `fl_sample_data`, comes from a real dataset of positive and unlabeled
cases, with one row per vessel-year and columns corresponding to vessel identity,
the features of the model based on AIS data (and briefly described in the section 
below), and columns to identify the positive cases (`known_offender` column), 
negative cases (`known_non_offender` column), and unlabeled cases (`0` values in 
`known_offender` and in `known_non_offender`). 
The original dataset contained more than 100 positive cases and almost
a million vessel-years of unlabeled data. The gear types included are
trawlers, longlines and squid jiggers. To generate `fl_sample_data`, we 
anonymized the vessels, kept most of the positives (127), and randomly sampled 10,000 
vessel-years from the unlabeled, so that there could be a manageable sample to 
run the functions in the package in a local computer. This sample does not 
contained real negative cases (the C188-certified vessels), as they were 
obtained under a non-disclosure agreement. If further negative cases are made 
available and it is possible to share them publicly, we could add them to the 
holdout dataset of the package in the future. 

In `fl_sample_data`, 100 of the positives are meant to be used to
train the model and the other 27 should be used as a holdout dataset. 9900 of the 
unlabeled would be used for training and the other 100 would be used as a 
holdout dataset. The holdout dataset is later used to compute performance metrics.
The rows with `holdout == 1` correspond to the holdout dataset and those with
`holdout == 0`, to the training dataset.

```{r setup, eval = F}
library(forcedlabor)
```


```{r load_data}
data("fl_sample_data")
training_df <- fl_sample_data |> filter(holdout == 0)
holdout_df <- fl_sample_data |> filter(holdout == 1)
```



## Model features

The features used for classification correspond to vessel characteristics and 
variables describing vessel activity at sea that can help identify vessel with
forced-labor-like movement patterns. 

The features for this model were computed from Automatic Identification System (AIS) data processed by Global Fishing Watch. 

The rationale behind why we selected these features is summarized below.

### Vessel characteristics

- *Gear type* shapes expected movement.
- *Length*, *tonnage* and *engine power* may enable longer, farther operations—factors 
associated with elevated forced-labor risk.

### Time at sea

Long voyages and infrequent port visits allow vessels to avoid inspections and limit
crew opportunities to report or exit abusive conditions.
Excessive overtime is a common indicator of abuse. The following variables are
included to capture unusually long daily or yearly effort:

- *number of AIS positions*
- *number of hours at sea*
- *number of voyages*
- *average voyage duration* (in hours)
- *total number of fishing hours* 
- *average daily fishing hours*

### Distance from shore and port

Operating far from shore may mean greater isolation and reduced oversight. The model includes: 
 
- *maximum distance from port* (km)
- *maximum distance from shore* (km)

### Multijurisdictional issues

Time in foreign EEZs or on the High Seas, and foreign port visits, increase
the jurisdictional complexity to address forced labor cases. The model includes: 

- *number of fishing hours in foreign EEZs*
- *number of fishing hours in the High Seas*
- *number of foreign port visits*

### Transshipment, encounters, and loitering

Encounters and loitering (stationary behavior at sea that may indicate a  
potential encounter with a non-AIS-broadcasting vessel) patterns can indicate
transshipment events. 
Transshipments can prolong the time at sea and enable crew transfers, both of
which are linked to higher risk of forced labor.
The model includes: 

- *number of encounters with other vessels at sea*
- *number of encounters with vessels reported to have engaged in forced labor*,
- *average encounter duration* (hours)
- *number of loitering events*
- *average loitering duration* (hours).

### AIS disabling events-"gaps"

Intentional AIS disabling can obscure activity and mask behaviors such as 
transshipment or other evasive practices. The model includes: 

- *number of intentional gaps*
- *average duration (in days) for these gaps* 
- *average distance* between the location where an intentional gap begins and the location where it ends
- *average distance from port during these gaps*
- *average distance from shore during these gaps* 


## Model workflow 

The `forcedlabor` package allows to execute the workflow described in [@joo_2023] (and updated in a paper in review).

```{r workflow, echo=FALSE, out.width="100%"}
here <- here::here()
img <- (glue::glue(here, "/vignettes/fig01_model_workflow.png"))
knitr::include_graphics(img, rel_path = FALSE)

```

- The function `fl_rfsetup()` sets up the model features, response variables,
and the parameters required to configure random forests (RFs): our model first 
fits a series of random forests, using the `ranger` engine within a `tidymodels` 
framework. 
- The function `fl_cvsetup()` takes the output of `fl_rfsetup()` and fixes
the number of folds (k), bags, seeds and downsampling ratio required to run the
cross-validations. These are described in the sections below.
- The function `fl_train()` fits the RFs for one seed and one bag using the 
`recipe` configuration from `fl_cvsetup()`, creating a
series of RF scores for the training data, and optionally, for the holdout
dataset. 
- The function `fl_classification()` implements the DEDPUL algorithm,
calculating a calibrated threshold `t` and classifying the average RF
scores against `t` to classify each of them as `positive` or `negative`. 
The calculation of confidence values can be done optionally. 
- The function `fl_perf_metrics()` calculates recall and specificity (if a
`holdout` dataset is provided). 



## Feature pre-processing and model configuration using `fl_rfsetup()` and `fl_cvsetup()`

Following the `tidymodels` framework, the function `fl_rfsetup()` calls `recipes` package functions to preprocess the training dataset, specifying roles and transformations for model features. 
Importantly, it distinguishes which features will be used as predictors, which one is the response variable and which one will be used as `control` variable to avoid data leakage. 

Vessel `MMSI` (Maritime Mobile Service Identity) --which can appear in several related cases--is thus used as a control variable against data leakage (all vessel-year cases from the same vessel were assigned to either the training or the holdout dataset). This corresponds to the column `ssvid`. 

Feature transformations included removing near-zero variance variables with `recipes::step_nzv()` and removing highly correlated variables (correlation value can be setup by parameter `corr_threshold`). 

`fl_rfsetup()` also calls the `parsnip` package to specify hyperparameter values of the RFs, such as the number of trees, the number of features that will be sampled at each split, the minimum node size and the regularization factor. 
The chosen model engine was the one in package `[ranger]`. 


```{r cv_splits}
features_model_characteristics <- c("gear",
                                    "engine_power_kw",
                                    "tonnage_gt",
                                    "length_m")
features_model_movement <- c("position_messages",
                             "hours",
                             "fishing_hours",
                             "average_daily_fishing_hours",
                             "fishing_hours_foreign_eez",
                             "fishing_hours_high_seas",
                             "max_distance_from_shore_km",
                             "max_distance_from_port_km",
                             "number_encounters",
                             "number_forced_labor_encounters",
                             "average_encounter_duration_hours",
                             "gaps_12_hours",
                             "average_off_distance_from_port_km",
                             "average_off_distance_from_shore_km",
                             "average_gap_days",
                             "average_gap_km",
                             "number_foreign_port_visits",
                             "number_loitering_events",
                             "average_loitering_duration_hours",
                             "average_voyage_duration_hours",
                             "number_voyages")

rf_setup <- fl_rfsetup(training_data = training_df,
                       y = "known_offender",
                       x = c(features_model_characteristics,
                             features_model_movement),
                       id = "indID",
                       dont_use = c("known_non_offender"),
                       control = "ssvid",
                       corr_threshold = 0.75,
                       rf_trees = 500,
                       rf_mtry = 5,
                       rf_min_n = 10,
                       rf_reg_factor = 1)
```

The function returns a list with the two elements corresponding to the setup above: 

- `$rf_recipe`, with the output from `recipes`
- `$rf_spec` with the RF specification by `parsnip`


### Function `fl_cvsetup()`

The function `fl_cvsetup()` creates the whole framework to fit the RF models for each seed, bag, and fold, from the training data, the model recipe and the RF specification. 
As shown in the Figure of the model workflow, we do a k-fold cross-validation to split
the data into analysis (to train an RF) and assessment (to generate RF scores)
for vessel-years that were not used to train the RF. 
In this case, we are doing a 5-fold cross-validation (that will use `ssvid` as a variable to control for data leakage),
so `num_folds = 5`.
In each analysis set, the unlabeled cases are more numerous than the positive cases.
For that reason, we downsample the unlabeled in the analysis set with a downsampling ratio.
In this case, we use a 1:1 ratio. The downsampling procedure is repeated `num_bags` times (in this case, 10 bags) for each of the five splits.
The partition of samples into analysis and assessment during cross-validation, 
the downsampling (`num_bags` times) and fitting the RFs for each bag and fold combination, 
is done for a given initial randomization seed. We repeat this process for `num_common_seeds = 3`.
With 10 bags and 3 seeds, this yields 30 RF scores per case (i.e. vessel-year). 


```{r fl_cvsetup}
num_folds <- 5
num_bags <- 10
down_sample_ratio <- 1
num_common_seeds <- 3

tictoc::tic()
cv_df <- fl_cvsetup(training_data = training_df,
                    fl_rec = rf_setup$rf_recipe,
                    rf_spec = rf_setup$rf_spec,
                    num_seeds = num_common_seeds,
                    num_bags = num_bags,
                    num_folds = num_folds,
                    down_sample_ratio = down_sample_ratio)
tictoc::toc()
```

The output of this function is a list with `seeds x bags` elements (here: `r num_common_seeds` seeds and `r num_bags` bags = `r num_common_seeds * num_bags` elements). 
Each element of this list includes the seed, bag number, recipe and a tibble object with the k-fold cross-validation splits, where the unlabeled data has been downsampled by `themis::step_downsample()` and  the folds have been created by `rsample::group_vfold_cv()`. 


## Training the models: function `fl_train()`

Once the training data has been prepped and the model recipe and specifications
set, we can now use function `fl_train()` to fit RF models and generate scores. 

Function `fl_train()` fits a RF for each fold for a bag and a seed configuration using
the previous recipe, specification steps and cross-validation setup.

- It applies a downsampling ratio of 1:1 to the training set
- This step returns a list with 2 elements: `train_probabilities` and 
`pred_probabilities` __if `holdout` was included__. `train_probabilities` is list
with *k* elements, which are tibbles. Each tibble contains RF scores 
for its corresponding assessment rows (from the k-fold validation). 
`pred_probabilities` is also a list with *k* tibbles. Each tibble contains RF scores
for the holdout rows, using the RF fitted with the analysis rows (from the k-fold validation).

The function is written to receive a single element of the output of 
`fl_cvsetup()` (`cv_df` in this example). So a single run (with 5 folds) 
could be generated like this:

```{r, eval = FALSE, warning=FALSE}
tictoc::tic()
one_trained_model <- fl_train(cv_setup = cv_df[[1]],
                              rf_spec = rf_setup$rf_spec,
                              holdout = holdout_df)
tictoc::toc()
```

Tibbles from both `train_probabilities` and `pred_probabilities` include the following columns: 
`.pred_1` (RF scores), `indID` (unique vessel identifier), `known offender`,
`known non offender` (unavailable in this dataset), and `holdout`.

To train RFs and get scores for all seeds and bags, we can use `for` loops
or `purrr::map()`, or parallelize in machines with very large RAM. 
In this example, let's use `purrr::map()`.

Note that the use of `holdout` is optional. If omitted, `fl_train()` will
produce RF scores for the training data only. If added, it will create
RF scores for new data that is not used in model training.

```{r fl_train, warning=FALSE}
tictoc::tic()
trained_models <- purrr::map(.x = cv_df,
                             ~fl_train(rf_spec = rf_setup$rf_spec,
                                       holdout = holdout_df))

tictoc::toc()

```

`trained_models` is composed of a list of length `bags x seeds` (in this
case, `r num_common_seeds * num_bags`).
Each element of that list contains the two elements described above, 
`train_probabilities` and `pred_probabilities` __if `holdout` was
included__.

Let's extract both lists and create a single dataframe for all

```{r train_proba_output}

train_probabilities_all <- purrr::map(trained_models, "train_probabilities") |> bind_rows()
pred_probabilities_all <- purrr::map(trained_models, "pred_probabilities") |> bind_rows()
prob_all <- bind_rows(pred_probabilities_all,
                      train_probabilities_all)
head(prob_all)
```

### Function `fl_classification()`: density-based algorithm for classification 
threshold

`fl_classification()` aggregates all RF scores for each row, 
estimates $\alpha$ (upper bound on positives in the unlabeled set)
and estimates a threshold `t` so that the proportion of 
predicted positives ≈ $\alpha$.

See [@ivanov_2020] for more details on the DEDPUL algorithm.


```{r classif_res, warning=FALSE}
tictoc::tic()
classif_res <- fl_classification(data = prob_all,
                                 steps = 1000,
                                 plotting = FALSE,
                                 filepath = NULL,
                                 threshold = seq(0, .99, by = 0.01),
                                 eps = 0.1,
                                 confidence_levels = TRUE
                                 )
tictoc::toc()
```

- Within `fl_classification()` we apply the calibrated
threshold `t` to each vessel’s mean RF score (`pred_mean`):  
  if `pred_mean > t`, then `pred_class = 1`, else `0`. This produces a binary prediction (positive or negative, stored in column `pred_class`). 
- The `confidence` value indicates how consistently a vessel’s
score falls on the same side of `t` across seeds × bags. Values near 1
indicate a very stable classification for that vessel-year (i.e. higher
confidence in the classification).

```{r classif_res_pred_conf}
class_pos <- sum(classif_res$pred_conf$pred_class == 1)
class_neg <- sum(classif_res$pred_conf$pred_class == 0)
```

In the case of this sample dataset, `r class_neg` vessel-years were classified as negative and `r class_pos` were classified as positive for forced labor risk. 

## Assessing model performance with function `fl_perf_metrics()`

The performance of the model is evaluated through:

- `recall`: the proportion of positive cases predicted by the model
as positive
- `specificity`: if negative holdout cases have been provided, it is computed as the
proportion of negative cases correctly predicted by the model as negatives

```{r perf_metrics}

tictoc::tic()
perf_metrics <- fl_perf_metrics(data = classif_res$pred_conf)
tictoc::toc()

print(perf_metrics)
```


### Fairness tests

Statistical non-discrimination or fairness measures aim to assess the absence or
degree of discrimination of given groups regarding classification. 

In this context, this meant assessing whether the model was equally successful 
(or unsuccessful) at classifying positive vessel-years across different fishing gears.


```{r fairness_gear}

# Recall by gear

gear_lookup <- training_df  |>
  distinct(indID, gear)

gear_lookup |> count(gear)

gear_lookup |> 
left_join(classif_res$pred_conf) |>
 group_by(gear) |> 
 mutate(
   known_offender = factor(as.character(known_offender), levels = c("0","1")),
   pred_class     = factor(as.character(pred_class),     levels = c("0","1"))
 ) |> 
 yardstick::recall(truth = known_offender,
        estimate = pred_class, event_level = "second")  |> 
 select(gear, .estimate)
```

Unsurprisingly, the randomized sample was not large enough for the model to learn to disentangle
forced-labor-like behaviors, but this was necessary to reduce computational costs for this vignette.





# References


