---
title: "Using the package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{using_the_package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)

devtools::load_all()
```

## Overview

`forcedlabor` is an R package to train a positive–unlabeled (PU) model to 
detect fishing vessels exhibiting behaviors or characteristics consistent 
with forced-labor offenders. 
Features (movement patterns and vessel characteristics) are computed from 
AIS data.
We have used Global Fishing Watch’s AIS data pipeline. 
The package assumes access to AIS (or other vessel tracking) data.

```{r setup, eval = F}
library(forcedlabor)
```

### Model workflow 

The model workflow contains the following:
Steps 1-3 involve data and model prep
Steps 4-8 use the package functions
Steps 9 and 10 are additional steps for assessing model performance and fairness
(not included in workflow plot)

```{r workflow, echo=FALSE, out.width="100%"}

img <- normalizePath(file.path("fig01_model_workflow.png"), mustWork = TRUE)
knitr::include_graphics(img, rel_path = FALSE)

```

The model workflow replicates that in [@joo_2023].

1. Data: the training data is loaded. 
This contains AIS-derived data at a yearly scale for positive vessels 
(`known_offender = 1`) and unlabeled vessels (`known_offender = 0`)
2. (Same) 5-fold CV: creates grouped folds by `source_id_number` so related 
vessels stay together, rotating the assessment fold per seed.
3. Fit random forests: for each fold × bag × seed, fit an RF using the 
recipe/spec, with 1:1 down-sampling applied to the analysis split.
4. Predicted scores: Collect out-of-fold probabilities `.pred_1` via 
package function `dev_ml_train()`
5. Predictions from all seeds: pool the `.pred_1` scores across all seeds and 
bags into `train_pred_proba`
6. Density-based threshold (DEDPUL): `ml_classification()` function averages 
scores to `pred_mean`, estimates alpha (α) on assessment data, and chooses a 
calibrated threshold (t)
7. Classification: Apply 't' to `pred_mean` to produce the binary label 
`pred_class` (1 positive, 0 negative)
8. Confidence: Compute a confidence score per vessel based on 
variation across seeds × bags.

Additional steps:

9. Model performance: Report recall on assessment folds and specificity 
if a holdout is provided using package function `ml_perf_metrics`
10. Fairness tests: Evaluate recall parity across non-modeled groups 
(e.g., flag_region, gear) as post-hoc diagnostics


## Pre-package data and model prep

Before using the package functions we load the training data
and prep the model

### 1. Data

#### Training data

The model starts with loading the required training data (`fl_training`).
This table contains AIS-derived data at a yearly scale for positive vessels 
(`known_offender = 1`) and unlabeled vessels (`known_offender = 0`). 


```{r load_data}

data("fl_training")

fl_training$known_non_offender <- as.factor(fl_training$known_non_offender)

levels(fl_training$known_non_offender) <- 
  c(levels(fl_training$known_non_offender),1)
```


##### Positives: offenders

The training dataset includes a list of positive (known_offender = 1) 
and unlabeled (known_offender = 0) vessels and their corresponding outputs 
from tracking data. 

Our offender data has come from external reports on forced labor. 
In many cases the exact date of the offence is unclear, hence why 
we work on a yearly scale. 
Reported cases were then matched to AIS data. 

##### (Optional) holdout dataset via `prediction_df`

- The model is not trained on negative cases (`known_non_offender = 1`)
- If negative data is available, it can be used in the `prediction_df` to 
calculate specificity (but `prediction_df` is not used in training)
- If non-offender data is not available, `prediction_df` should be set to `NULL` 
    - Note that when `prediction_df` = `NULL`, specificity cannot be computed


##### Unlabeled

All fishing vessels with no reported forced labor offences. 


##### Predictors

We use 26 variables as predictors. 
The features used for classification corresponded to vessel characteristics 
and movement patterns computed from Automatic Identification System (AIS) data.

The rationale behind why we selected these features is provided below.

**Vessel characteristics**

Gear type shape expected movement.
Length, tonnage and engine power may enable longer, farther operations—factors 
associated with elevated forced-labor risk.

**Fishing hours**

Excessive work is a common indicator of abuse, so we include total and average 
daily fishing hours to capture unusually long daily or yearly effort.

**Distant water fishing**

Time in foreign EEZs or on the High Seas and large distances from shore and 
port increase jurisdictional complexity and remoteness, raising risk of forced
labor.

**Transshipment, encounters, and loitering**

Encounter and loitering patterns can indicate transshipment. 
Transhipment can prolong the time at sea, and enabling crew transfers.
Both of which are linked to higher risk of forced labor.

**AIS gaps**

Gaps in AIS can obscure activity and mask behaviours such as transshipment or 
other evasive practices.

**Foreign port visits**

Frequent foreign port calls can signal distant-water operations or recruitment 
pathways via ports with weaker oversight.

**Length of time at sea**

Long voyages and infrequent port returns reduce monitoring and limit crew 
opportunities to report or exit abusive conditions.

```{r training_data}
dplyr::glimpse(fl_training)
```

### 2. (Same) 5-fold Cross Validation

Once the training data is loaded, we set up our 5-fold CV.
Here, the data (positives and unlabeled) are split into five groups of 
roughly equal size.
One of the groups is designated as the test set, and the others are part of 
the training set. 
The role of the test set is rotated among the group five times.

To reduce the risk of data leakage, we keep data from the same source (in our
case from the same forced labor report) in the same group.
This means that vessels with the same `source_id_number` will either all be
in training or all in test.
We do this as two vessels from the same report are likely to have similar
behaviors and characteristics, and so so when using one of them in the 
training set, the model would likely
perform well on the other one, an artificial effect of using related cases to 
train and test.

#### Cross-validation setup (folds, bags, seeds)

- `num_folds <- 5` splits the data into 5 groups of roughly equal size
    - More folds means better use of data and therefore lower likelihood of 
    bias, but also means more computational costs
- `num_bags` sets the number of bags to run as 5
  - More bags mean more stable probability estimates across resampling
- `down_sample_ratio` downsampling ratio set to 1:1
  - To reduce the weight of the unlabeled cases in the model, we randomly 
  downsampled them in the training set with a 1-1 ratio, i.e. the number of 
  positive and unlabeled cases used for training would be equal
  - This downsampling ratio is used later by the `dev_ml_train` (Step 4)
  but is included here for clarity
- `num_common_seeds <- 3` sets 3 random seeds 
  - Setting up common seeds to use anywhere that uses random numbers.
  - Using fewer seeds is faster, but will be more sensitive to random chance
- `cv_splits_all` checks that we don't have any splitting across source_id 
for the analysis and assessment datasets


```{r cv_splits}

num_folds <- 5 #5
num_bags <- 2 #5 
down_sample_ratio <- 1
num_common_seeds <- 2 #3 
```


### 3. Fit random forests

#### Writing the recipe

Next, we want to write the recipe for the model. 
Here we set the roles of the different variables that will be used. 

**Setting the roles of the variables used**

- `known_offender` is set as the **outcome** i.e. the target we predict
- `indID` is set as an "id" value and is not used as a predictor
- `known_non_offender` is not used as a predictor (role "dont_use")
    - if available, use `known_non_offender`  in a holdout `prediction_df` to 
      compute specificity later
- `source_id_number` is set as "control"
- `flag_region` is not used as a predictor (role "dont_use")
      - it is reserved for post-hoc fairness checks 
- **Predictors**: All remaining variables in `fl_training` are treated as 
predictors

**Variable preprocessing**

- We use `step_nzv()` to retain predictors with non-zero variance
  - This step diagnoses predictors that have one unique value 
  (i.e. are zero variance predictors) or predictors that have both of the 
  following characteristics:
    1. they have very few unique values relative to the number of samples
    2. the ratio of the frequency of the most common value to the frequency of 
  the second most common value is large.
- Using `step_corr` we only keep predictors with bivariate correlations smaller 
than 0.75.


```{r fl_rec}

fl_rec <- recipes::recipe(known_offender ~ .,
                          data = fl_training) |>
  recipes::update_role(indID,
                       new_role = "id") |>
  recipes::update_role(flag_region, known_non_offender,
                       new_role = "dont_use")  |>
  recipes::update_role(source_id_number, new_role = "control")  |>
  recipes::step_nzv(recipes::all_predictors())  |>
  recipes::step_corr(recipes::all_numeric(), threshold = 0.75)

```


#### Specifying the model

Next, we define the classifier. 
We use a Random Forest via `parsnip` with the `ranger` engine.

- The type of model used is a random forest `rand_forest()`
- `trees = 500` sets the number of trees contained in the ensemble is set to 
500
  - The larger the number of trees, the more stable the predictions - but with
  a higher computational cost
- `mtry = 1` sets the number of features that will be sampled at each
split
  -  `mtry` must be ≤ the number of predictors in the dataset
  - Lower values encourage diversity between trees (higher variance, 
  less correlation)
  - Higher values let the model consider more predictors, potentially improving 
  accuracy but also with a risk of overfitting
- `min_n = 15` sets the minimum node size (i.e. the number of vessel-years that 
a final node should have)
  - Larger values create less complex trees, and may result in underfitting
  - Smaller values create more complex trees, and may result in overfitting
- `set_mode("classification")` the prediction outcome mode is set as 
"classification" 
- `set_engine("ranger", regularization.factor = 0.5)` 
- `set_engine()` 
  - `ranger` is particularly useful when dealing with data with a high number of
  features as it provides memory-efficient yet quick implementation
- `regularization.factor = 0.5` sets the regularization factor works by 
penalizing new variables by multiplying the splitting criterion
by a factor
  - As a result, the tree is biased to keep splitting on features it already 
  trusts, rather than constantly introducing fresh ones
    - `f = 0.5`: half the gain for “new” variables, forcing the forest to only 
    bring in a new predictor if it improves impurity by at least twice as much
    the default would require


```{r fl_spec}

rf_spec <-
  parsnip::rand_forest(trees = 500,
                       mtry = 1,
                       min_n = 15) |>
  parsnip::set_mode("classification") |>
  parsnip::set_engine("ranger", regularization.factor = 0.5)

```

#### Parallelization strategy

We parallelize the bag x seed training across local CPU cores using `future`

- `parallel_plan`:
  - "multisession" spawns separate R sessions (safe on Windows & RStudio).
  - "multicore" uses Unix forking (Linux/Mac in plain R; not on Windows/RStudio).
  - "psock" is a portable fallback if multisession misbehaves (slightly slower).
- `free_cores = 4`: keep 4 cores free for your system
- `future.globals.maxSize`: max bytes of objects sent to workers 

```{r parallel_plan}
## parallelization strategy
parallel_plan <- "multisession" # multisession if running from RStudio, or
# multicore if from Linux, Mac and plain R, or
# psock if multisession is not working well and you need to try something else
if (parallel_plan == "multisession") {
  utils::globalVariables("multisession")
}
free_cores <- 4 # add more if you need to do many things at the same time
oopts <- options(future.globals.maxSize = 10200*1024^2)  ## 15 GB
# options(future.globals.maxSize = 50 * 1024 ^ 3)


```


## Using the `forcedlabor` package functions

Once the training data has been prepped and the model recipe and specifications
set, we can now use the package functions to fit models and generate scores. 

### 4. Predicted scores (using `dev_ml_train`)

- `dev_ml_train`: fits a Random Forest for each fold x bag x seed using
the previous recipe and specification steps
- It applies a downsampling ratio of 1:1 to the training set
- This step returns `train_pred_proba` a table with probabilities for
assessment rows 

### 5. Predictions from all seeds (using `dev_ml_train`)

- `dev_ml_train` also pools scores across all seeds and bags


```{r train_pred_proba}

tictoc::tic()
cv_df<- dev_cv_setup(training_data = fl_training,
                       num_folds = num_folds,
                       num_bags = num_bags,
                       num_seeds = down_sample_ratio,
                       fl_rec = fl_rec,
                       rf_spec = rf_spec,
                       down_sample_ratio = down_sample_ratio,
                       tune = FALSE,
                       grid = NULL)
trained_models <- dev_ml_train(cv_setup = cv_df,
                           free_cores = free_cores,
                           parallel_plan = parallel_plan,
                           save_dir = NULL)#"../models/test")
train_pred_proba = trained_models$train_probabilities
tictoc::toc() 

### training and prediction (scores)

#tictoc::tic()
#train_pred_proba <- ml_train_predict(
#  fl_rec = fl_rec,
#  rf_spec = rf_spec,
#  cv_splits_all = cv_splits_all,
#  bag_runs = bag_runs,
#  down_sample_ratio = down_sample_ratio,
#  parallel_plan = parallel_plan,
#  free_cores = free_cores,
#  prediction_df = NULL
#)
#tictoc::toc()


#ml_train(fl_rec = fl_rec, 
#         rf_spec = rf_spec,
#         cv_splits_all = cv_splits_all,
#         bag_runs = bag_runs,
#         down_sample_ratio = down_sample_ratio,
#         parallel_plan = parallel_plan,
#         free_cores = free_cores,
#         prediction_df = NULL,
#         save_dir = "../models")
  
```

`train_pred_proba` contains the following:

- `common_seed`: which randomization seed was used to generate both your 
down‑sampling and your CV splits for this batch of runs
- `bag`: which down‑sample “bag” this row belongs to
- `counter`: a simple row index for the combination of seed x bag grid
- `id`: the fold identifier from your CV split
- `.pred_1`: the model’s predicted probability that this particular row is 
positive
- `indID`: a unique vessel identifier
- `known offender`: if the vessel is positive
- `known non offender`: if the vessel is negative
- `holdout`: if 0, one of your CV assessment folds, if 1, an external 
hold‑out dataset (if passed via prediction_df)
    - If you have no `prediction_df = NULL`, all rows have `holdout == 0`


```{r train_pred_proba_output}

dplyr::glimpse(train_pred_proba)

```

### 6. Loading model and predicting over new data 

Previously trained models can be loaded from your local disk using `dev_ml_load` based on your cross-validation setup (output of `dev_cv_setup`) and used to predict over new datasets using `dev_ml_predict`. 

```{r}
loaded_models <- dev_ml_load(cv_setup = cv_df,
                             save_dir = "../models/test")
#predictions = dev_ml_predict(trained_models = loaded_models,
#                             new_data = fl_predict,
#                             free_cores = free_cores,
#                             parallel_plan = parallel_plan)
```

### 7. Density-based algorithm for classification threshold (using `ml_classification`)

`ml_classification` aggregates `.pred_1` per `indID` to `pred_mean`, 
estimates α (upper bound on positives in the unlabeled set), 
and chooses a calibrated threshold `t` so the share of 
predicted positives ≈ α (DEDPUL).

See [@ivanov_2020] for more details on the DEDPUL alogirthm. 

**Tunable args:**

- `steps`: when setting `steps` there is a balance between too few steps 
overpredicting positives, and too many slowing down for little gain
  - `steps` must be at least 3 in order to calculate the second derivative
- `threshold` and `eps`: these are used to set the threshold for what is
considered positive or negative
  - `threshold` sets the list of candidate cut‑offs to see what gets us 
  closest to the alpha, so that the threshold of classification t is such that 
  the proportion of positive predictions is equal (or approximately equal) to α
  - `eps` is the range from this alpha that we are willing to accept
    - setting `eps` too low will likely return an error, as none of the 
    candidates
are within the range
- If `plotting = TRUE`, a D vs. alpha plot is generated


```{r classif_res, warning=FALSE}

#tictoc::tic()
#classif_res <- ml_classification(data = train_pred_proba,
#                                 steps = 1000,
#                                 plotting = FALSE,
#                                 filepath = NULL,
#                                 threshold = seq(0, .99, by = 0.01),
#                                 eps = 0.01,
#                                 parallel_plan = parallel_plan,
#                                 free_cores = free_cores)
#tictoc::toc()


```


### 8. Classification (using `ml_classification`)

- Within `ml_classification` we apply the calibrated threshold `t` to 
each vessel’s mean score:  
  `pred_class = 1` if `pred_mean > t`, else `0`  
- This produces the binary `pred_class`


### 9. Confidence for each vessel year classification (using `ml_classification`)

- The `confidence` value indicates how consistently a vessel’s 
score falls on the same side of `t` across seeds × bags
- Computed only when multiple runs exist
- Values near 1 indicate very stable 
classification.


```{r classif_res_pred_conf}

#dplyr::glimpse(classif_res$pred_conf)

```

## Additional steps: Performance and Fairness


### 10. Performance: Recall and specificity (using `ml_perf_metrics`)

The performance of the model is evaluated through:

- `recall`: the proportion of positive cases predicted by the model
as positive
- `specificity`: if a `prediction_df` has been provided, the
proportion of  negative cases predicted by the model as negatives

```{r perf_metrics}

#tictoc::tic()
#perf_metrics <- forcedlabor::ml_perf_metrics(data = classif_res$pred_conf)
#tictoc::toc()

#print(perf_metrics)


```


### 11. Fairness tests

Statistical non-discrimination or fairness measures aim to assess the absence or
degree of discrimination of given groups regarding classification. 

We evaluated the model’s fairness as its capacity to correctly identify reported 
positive cases by flag region and fishing gear.

The first example below shows how to assess fairness based on 'flag region'

- The fairness test assesses whether the model is equally successful at 
classifying offenders as positive (recall) when considering different flag 
regions
  - Flags could be a sensitive characteristic, and potentially associated with
  discrimination
  - As we did not have enough cases to assess based on each individual Flag 
  State, we compute and compare fairness for two groups: "Asian" and "Other"


```{r fairness_flag}

# Recall by flag_region

#region_lookup <- fl_training %>%
#  distinct(indID, flag_region)

#classif_res$pred_conf %>%
#  left_join(region_lookup, by = "indID") %>%
#  group_by(flag_region) %>%
#  mutate(
#    known_offender = factor(as.character(known_offender), levels = c("0","1")),
#    pred_class     = factor(as.character(pred_class),     levels = c("0","1"))
#  ) %>%
#  recall(truth = known_offender, 
#         estimate = pred_class, event_level = "second") %>%
#  select(flag_region, .estimate)
```



# References


